---
title: "ANPC MRMS_exploreR v1"
output:
  pdf_document: default
  html_document: default
    df_print: paged
---
This notebook is designed for use with the ANPC MRMS data. 
Section 1: explore MRMS data and visualise data quality.

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

Section 1 - MRMS exploreR

This notebook is designed to explore the MRMS dataset and perform signal drift correction and QC check it. The script generates a report and a final dataset that can be used for data modeling.

1. prepare project: 
  -> load packages and functions

```{r, mrms_exploreR packages and functions,  eval = FALSE, echo = FALSE, results = FALSE, warning=FALSE, message=FALSE}
# reset environment
if(length(names(sessionInfo()$otherPkgs)) > 0){
invisible(lapply(paste0('package:',names(sessionInfo()$otherPkgs)),detach,character.only=TRUE, unload=TRUE))
}

# load packages
package_list <- c("plyr", "tidyverse", "janitor", "gridExtra", "ggpubr", "readxl", "cowplot", "scales", "stats", "devtools", "metabom8", "shiny", "plotly", "svDialogs", "DataEditR", "htmlwidgets", "httr", "htmlTable", 'statTarget', 'knitr', 'mzR')
loaded_packages <- lapply(package_list, require, character.only = TRUE)
rm(loaded_packages, package_list)

#welcome message
dlg_message("Welcome to MRMS exploreR! :-)", type = 'ok')

# set parent github URL
parent_github_url <- "https://raw.githubusercontent.com/lukewhiley/mrms_exploreR/main"

# lipid_exploreR url - for downloading functions

parent_github_function_url <- "https://raw.githubusercontent.com/lukewhiley/metabolomics_code/main"

# load custom functions from github

#PCA function
PCA_QC_function <- GET(url = paste0(parent_github_function_url, "/functions/2021-LGW-universal_PCA_function.R")) %>% 
                                    content(as = "text")
eval(parse(text = PCA_QC_function), envir = .GlobalEnv)
rm(PCA_QC_function)


# download and parse data preparation script
mrms_exploreR_script <- GET(url = paste0(parent_github_url, "/scripts/2021-LGW-mrms_exploreR_script.R")) %>% 
                               content(as = "text")
eval(parse(text = mrms_exploreR_script), envir = .GlobalEnv)
rm(mrms_exploreR_script)

print(paste0("Succesfully imported ", project_name, ": ", nrow(mrms_exploreR_data$data_unprocessed), " samples; ", length(mrms_exploreR_data$feature), " MRMS features"))
print(paste0("There are ", mrms_exploreR_data$data_unprocessed$batch %>% unique() %>% length(), " batches in the dataset"))
print(paste0("There are ", mrms_exploreR_data$data_unprocessed$plateID %>% unique() %>% length(), " plates in the dataset"))
print(paste0("There are a total of ", total_data_points, " data points"))
print(paste0(total_percentage_zero_values, " % of total data points are missing or are a 0 value"))


```

MRMS feature heatmap
  -> to observe missing value distribution
  
```{r, mrms heatmap,  eval = FALSE, echo = FALSE, results = FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=3}

mrms_heatmap


```


QC check 1:
  -> Sums all features and plots output
  -> helps visualise outlier samples/mis-injections
  
```{r, summed_TIC_checkR,  eval = FALSE, echo = FALSE, results = FALSE, warning=FALSE, message=FALSE}
setwd(project_dir)

mrms_exploreR_data$data_for_TIC_QC_check <- mrms_exploreR_data$data_unprocessed

summed_TIC_checkR_script <- GET(url = paste0(parent_github_url, "/scripts/2021-LGW-mrms-summed_TIC_checkR_script.r")) %>% 
  content(as = "text")
eval(parse(text = summed_TIC_checkR_script), envir = .GlobalEnv)
rm(summed_TIC_checkR_script)

if(temp_answer == "all"){mrms_exploreR_data$data_tic_filtered <- mrms_exploreR_data$data_for_TIC_QC_check %>% filter(!sampleID %in% tic_qc_fail$sampleID)}
if(temp_answer == "samples"){mrms_exploreR_data$data_tic_filtered <- mrms_exploreR_data$data_for_TIC_QC_check %>% filter(!sampleID %in% tic_qc_fail_samples$sampleID)}
if(temp_answer == paste0(qc_type)){mrms_exploreR_data$data_tic_filtered <- mrms_exploreR_data$data_for_TIC_QC_check %>% filter(!sampleID %in% tic_qc_fail_ltr$sampleID)}
if(temp_answer == "none"){mrms_exploreR_data$data_tic_filtered <- mrms_exploreR_data$data_for_TIC_QC_check}

kable(tic_qc_fail,
      caption = paste0(nrow(tic_qc_fail), " samples FAILED the TIC QC check  ", nrow(tic_qc_fail_ltr),"  were QCs."))

tic_check_p_nc <- tic_check_p; tic_check_p_nc

temp_answer_tic_nc <- temp_answer_tic
print(paste0("Samples were considered outliers if their TIC was outside ", temp_answer_tic_nc, " % of the median "))
print(paste0(nrow(tic_qc_fail), " samples FAILED the TIC QC check  ", nrow(tic_qc_fail_ltr)," of which were QCs."))
print(paste0("The dataset now contains ", nrow(mrms_exploreR_data$data_tic_filtered), " samples"))

```



QC check 2:
  -> This QC check performs 2 basic feature QC checks
  -> 1: number of zero values in each feature
  -> 2: % RSD of each feature in the QC of choice
  
```{r, mrms_exploreR signal QC check,  eval = FALSE, echo = FALSE, results = FALSE, warning=FALSE, message=FALSE}

mrms_exploreR_data$data_for_signal_QC_check <- mrms_exploreR_data$data_tic_filtered

# download and parse data preparation script
mrms_exploreR_qc_check_1 <- GET(url = paste0(parent_github_url, "/scripts/2021-LGW-mrms_exploreR_QC_checkeR.R")) %>% 
                               content(as = "text")
eval(parse(text = mrms_exploreR_qc_check_1), envir = .GlobalEnv)
rm(mrms_exploreR_qc_check_1)



if(remove_qc == "keep"){mrms_exploreR_data$data_tic_filtered_qc_filtered <- mrms_exploreR_data$data_for_signal_QC_check}
if(remove_qc == "remove"){mrms_exploreR_data$data_tic_filtered_qc_filtered <- mrms_exploreR_data$data_for_signal_QC_check %>% 
  select(sampleID, type, run_order, plateID, all_of(metabolite_list_filtered_2))}

percentage_of_zero_plot_data_nc <- percentage_of_zero_plot_data
rsd_plot_data_nc <- rsd_plot_data

QC_p_1_nc <- QC_p_1; QC_p_1_nc
QC_p_2_nc <- QC_p_2; QC_p_2_nc

paste0("QC check 1: ", nrow(percentage_of_zero_plot_data_pass), " passed QC check and had < 50% zero values.  ", nrow(percentage_of_zero_plot_data_fail), " failed QC check and had > 50% zero values")
paste0("QC check 2: ", nrow(rsd_plot_data_pass), " features had an % RSD of < 30% in replicate QC samples.  ", nrow(rsd_plot_data_fail), " features had an RSD > 30% in the PQC.")


```
  

Quick PCA

```{r, PCA 1,  eval = FALSE, echo = FALSE, results = FALSE, warning=FALSE, message=FALSE}



QC_PCA_1_nc <- lgw_pca(FUNC_individual_multivariate_data = mrms_exploreR_data$data_tic_filtered_qc_filtered,
        FUNC_metabolite_list = mrms_exploreR_data$feature,
        FUNC_colour_by = "type",
        FUNC_plot_label = "sampleID",
        FUNC_scaling = "Pareto"); QC_PCA_1_nc

QC_PCA_2_nc <- lgw_pca(FUNC_individual_multivariate_data = mrms_exploreR_data$data_tic_filtered_qc_filtered,
        FUNC_metabolite_list = mrms_exploreR_data$feature,
        FUNC_colour_by = "plateID",
        FUNC_plot_label = "sampleID",
        FUNC_scaling = "Pareto"); QC_PCA_2_nc

```


Signal drift correction using the statTarget package:
  -> Performs mass spectrometry signal intensity drift correction
  -> user has the choice of Random Forrest (RF) or Loess correction
  -> signal drift images are output into a folder in the project directory
  
```{r, signal drift checkR script,  eval = FALSE, echo = FALSE, results = FALSE, warning=FALSE, message=FALSE}
setwd(project_dir)

mrms_exploreR_data$data_for_signal_drift_correction <- mrms_exploreR_data$data_for_signal_QC_check

signal_drift_choice <- "blank"
while(signal_drift_choice != "yes"& signal_drift_choice != "no"){
  signal_drift_choice <- dlgInput("Do you want to use a signal drift correction?", "yes/no")$res
}

if(signal_drift_choice == "yes"){
  
signal_drift_correct_script <- GET(url = paste0(parent_github_url, "/scripts/2021-LGW-mrms-signal_driftR_script.r")) %>% 
                                     content(as = "text")
eval(parse(text = signal_drift_correct_script), envir = .GlobalEnv)

mrms_exploreR_data$data_tic_filtered_qc_filtered_signal_drift <- corrected_data
  }

```

Re-perform the above plots on the signal corrected data

SIGNAL CORRECTED QC check 1:
  -> Sums all features and plots output
  -> helps visualise outlier samples/mis-injections
  
```{r, CORRECTED summed_TIC_checkR,  eval = FALSE, echo = FALSE, results = FALSE, warning=FALSE, message=FALSE}
setwd(project_dir)

mrms_exploreR_data$data_for_TIC_QC_check <- mrms_exploreR_data$data_tic_filtered_qc_filtered_signal_drift

summed_TIC_checkR_script <- GET(url = paste0(parent_github_url, "/scripts/2021-LGW-mrms-summed_TIC_checkR_script.r")) %>% 
  content(as = "text")
eval(parse(text = summed_TIC_checkR_script), envir = .GlobalEnv)
rm(summed_TIC_checkR_script)

if(temp_answer == "all"){mrms_exploreR_data$corrected_data_tic_filtered <- mrms_exploreR_data$data_for_TIC_QC_check %>% filter(!sampleID %in% tic_qc_fail$sampleID)}
if(temp_answer == "samples"){mrms_exploreR_data$corrected_data_tic_filtered <- mrms_exploreR_data$data_for_TIC_QC_check %>% filter(!sampleID %in% tic_qc_fail_samples$sampleID)}
if(temp_answer == paste0(qc_type)){mrms_exploreR_data$corrected_data_tic_filtered <- mrms_exploreR_data$data_for_TIC_QC_check %>% filter(!sampleID %in% tic_qc_fail_ltr$sampleID)}
if(temp_answer == "none"){mrms_exploreR_data$corrected_data_tic_filtered <- mrms_exploreR_data$data_for_TIC_QC_check}

tic_check_p_c <- tic_check_p; tic_check_p_c

kable(tic_qc_fail,
      caption = paste0(nrow(tic_qc_fail), " samples FAILED the TIC QC check  ", nrow(tic_qc_fail_ltr),"  were QCs."))

temp_answer_tic_cc <- temp_answer_tic
print(paste0("Samples were considered outliers if their TIC was outside ", temp_answer_tic_c, " % of the median "))
print(paste0("The dataset now contains ", nrow(mrms_exploreR_data$data_tic_filtered), " samples"))

```



SIGNAL CORRECTED QC check 2:
  -> This QC check performs 2 basic feature QC checks
  -> 1: number of zero values in each feature
  -> 2: % RSD of each feature in the QC of choice
  
```{r, mrms_exploreR signal QC check,  eval = FALSE, echo = FALSE, results = FALSE, warning=FALSE, message=FALSE}

mrms_exploreR_data$data_for_signal_QC_check <- mrms_exploreR_data$corrected_data_tic_filtered

# download and parse data preparation script
mrms_exploreR_qc_check_1 <- GET(url = paste0(parent_github_url, "/scripts/2021-LGW-mrms_exploreR_QC_checkeR.R")) %>% 
                               content(as = "text")
eval(parse(text = mrms_exploreR_qc_check_1), envir = .GlobalEnv)
rm(mrms_exploreR_qc_check_1)



if(remove_qc == "keep"){mrms_exploreR_data$corrected_data_tic_filtered_qc_filtered <- mrms_exploreR_data$data_for_signal_QC_check}
if(remove_qc == "remove"){mrms_exploreR_data$corrected_data_tic_filtered_qc_filtered <- mrms_exploreR_data$data_for_signal_QC_check %>% 
  select(sampleID, type, run_order, plateID, all_of(metabolite_list_filtered_2))}


QC_p_1_c <- QC_p_1
QC_p_2_c <- QC_p_2; QC_p_2_c

paste0("QC check 1: ", nrow(percentage_of_zero_plot_data_pass), " features had an % RSD of < 30% in replicate QC samples.  ", nrow(percentage_of_zero_plot_data_fail), " features had an RSD > 30% in the PQC.")
paste0("QC check 2: ", nrow(rsd_plot_data_pass), " features had an % RSD of < 30% in replicate QC samples.  ", nrow(rsd_plot_data_fail), " features had an RSD > 30% in the PQC.")


```
  

SIGNAL CORRECTED Quick PCA

```{r, CORRECTED PCA 1,  eval = FALSE, echo = FALSE, results = FALSE, warning=FALSE, message=FALSE}

QC_PCA_1_c <- lgw_pca(FUNC_individual_multivariate_data = mrms_exploreR_data$corrected_data_tic_filtered_qc_filtered,
        FUNC_metabolite_list = mrms_exploreR_data$feature,
        FUNC_colour_by = "type",
        FUNC_plot_label = "sampleID",
        FUNC_scaling = "Pareto"); QC_PCA_1_c

QC_PCA_2_c <- lgw_pca(FUNC_individual_multivariate_data = mrms_exploreR_data$corrected_data_tic_filtered_qc_filtered,
        FUNC_metabolite_list = mrms_exploreR_data$feature,
        FUNC_colour_by = "plateID",
        FUNC_plot_label = "sampleID",
        FUNC_scaling = "Pareto"); QC_PCA_2_c

```




























Produce the final HTML QC report
  -> script produces a final report that is interactive and can be used to evaluate the quality of the data
  
```{r, lipid_exploreR report export,  eval = FALSE, echo = FALSE, results = FALSE, warning=FALSE, message=FALSE}
setwd(project_dir)
# produce a final report

if(signal_drift_choice == "yes"){
fileConn<-file(paste0(project_dir_html, "/ANPC_mrms-report-generatoR.r"))
writeLines(GET(url = paste0(parent_github_url, "/scripts/2021-LGW-mrms-report_generatoR_script_v2.r")) %>%
                 content(as = "text"), fileConn)
close(fileConn)
}

if(signal_drift_choice == "no"){
fileConn<-file(paste0(project_dir_html, "/ANPC_mrms-report-generatoR.r"))
writeLines(GET(url = paste0(parent_github_url, "/scripts/2021-LGW-mrms-no_signal_drift_report_generatoR_script_v2.R")) %>%
                 content(as = "text"), fileConn)
close(fileConn)
}

rmarkdown::render(input = paste0(project_dir_html, "/ANPC_mrms-report-generatoR.r"),
                  output_format = "html_document",
                  output_dir = paste(project_dir_html),
                  output_file = paste0(project_name, "_", user_name, "_mrms_exploreR_QC_report.html")
                  )

file.remove(paste(paste0(project_dir_html, "/ANPC_mrms-report-generatoR.r")))

browseURL(paste0(project_dir_html, "/",project_name, "_", user_name, "_mrms_exploreR_QC_report.html")) #open plotly widget in internet browser

# if(signal_drift_choice == "no"){
# write_csv(lipid_exploreR_data[["individual_lipid_data_sil_tic_intensity_filtered_ratio_rsd"]], 
#           paste0(project_dir, "/", Sys.Date(), "_", project_name, "_", user_name, "_lipidexploreRv2_individual_lipid_data.csv")
#           )
# 
# write_csv(lipid_exploreR_data[["class_lipid_data_sil_tic_intensity_filtered_ratio_rsd"]], 
#           paste0(project_dir, "/", Sys.Date(), "_", project_name, "_", user_name, "_lipidexploreRv2_summed_lipid_class_data.csv")
#           )
# }
# 
# if(signal_drift_choice == "yes"){
# write_csv(lipid_exploreR_data[["individual_lipid_data_sil_tic_intensity_filtered_ratio_signal_drift"]], 
#           paste0(project_dir, "/", Sys.Date(), "_", project_name, "_", user_name, "_lipidexploreRv2_individual_lipid_data.csv")
#           )
# 
# write_csv(lipid_exploreR_data[["class_lipid_data_sil_tic_intensity_filtered_ratio_signal_drift"]], 
#           paste0(project_dir, "/", Sys.Date(), "_", project_name, "_", user_name, "_lipidexploreRv2_summed_lipid_class_data.csv")
#           )
# }

dlg_message("Lipidomics pre-processing and QC complete. Thanks for using skylineR and lipid exploreR. TTFN.", type = 'ok')
dlg_message("PS. an official html report can be found in the project folder", type = 'ok')

```










